---
title: "Chapter 1"
author: "Brian A. Fannin"
output: html_document
---

Notes and scribbles while reading Introductory Time Series with R by Cowpertwait and Metcalfe.

We start by learning a bit about the `ts` object in R and also about the basics of time series. A time series is an ordered sequence of values. The `ts` object will store metadata about the `start` and `end` times, as well as the `frequency` of observations.

`ts` is an S3 object, with 20 generic functions implemented. Weirdly, it has no names. Using `as.list` will simply return a list whose elements match a vector of the observations.

```{r }
data(AirPassengers)
AP <- AirPassengers

class(AP)

start(AP)
end(AP)
frequency(AP)

methods(class = 'ts')

names(AP)

mojo <- as.list(AP)

str(AP)
plot(AP)

plot(aggregate(AP))
```

`cycle` and `frequency` are two fairly relevant bits of metadata. For our first sample data, we know that the frequency is 12, which means 12 observations per cycle.

```{r }
cycle(AP)

boxplot(AP ~ cycle(AP))
```

Because the frequency of this time series is set to be 12, we can use the `cycle` function to produce that nice boxplot.

Can we change the frequency? Let's try.


```{r error=TRUE}
mojo <- AP
cycle(mojo) <- 4
frequency(mojo) <- 4
```

Hmm, that doesn't do it. Looks as though we have to set that at construction. Before we do, though, let's have a look at their construction example.

```{r error=TRUE}
URL <- "http://www.massey.ac.nz/~pscowper/ts/Maine.dat"
Maine.month <- read.table(URL, header=TRUE)
```

Data isn't available. Bummer. I even went to the university website and nothing doing. Grr.

OK, the important thing is that `frequency` is used to subdivide an interval into smaller increments which (probably?) relate to one another and form seasonal patterns. Probably.

Let's have another look at `aggregate` as that's how C&M convert a monthly time series into an annual one.
OK, having done that, I don't think that's the way to go. `aggregate` adds successive values together like `cumsum`, but the difference is that it pays attention to the `frequency`. When we aggregate the airline passenger data, it will collapse everything to one year. If we truly want quarterly data, we have to approach this differently.

```{r }
aggAP <- aggregate(AP)
class(aggAP)

length(aggAP)
length(AP)

(length(AP) / frequency(AP)) == length(aggAP)
```


There may be a `ts` way to do this, but I'm going to cheat and explore how to do that using `dplyr` or some such thing.

Tomorrow.