---
title: "Chapter 1"
author: "Brian A. Fannin"
output: html_document
---

Notes and scribbles while reading Introductory Time Series with R by Cowpertwait and Metcalfe.

We start by learning a bit about the `ts` object in R and also about the basics of time series. A time series is an ordered sequence of values. The `ts` object will store metadata about the `start` and `end` times, as well as the `frequency` of observations.

`ts` is an S3 object, with 20 generic functions implemented. Weirdly, it has no names. Using `as.list` will simply return a list whose elements match a vector of the observations.

```{r }
data(AirPassengers)
AP <- AirPassengers
rm(AirPassengers)

class(AP)

start(AP)
end(AP)
frequency(AP)

methods(class = 'ts')

names(AP)

mojo <- as.list(AP)

str(AP)
plot(AP)

plot(aggregate(AP))
```

`cycle` and `frequency` are two fairly relevant bits of metadata. For our first sample data, we know that the frequency is 12, which means 12 observations per cycle.

```{r }
cycle(AP)

boxplot(AP ~ cycle(AP))
```

Because the frequency of this time series is set to be 12, we can use the `cycle` function to produce that nice boxplot.

Can we change the frequency? Let's try.


```{r error=TRUE}
mojo <- AP
cycle(mojo) <- 4
frequency(mojo) <- 4
```

Hmm, that doesn't do it. Looks as though we have to set that at construction. Before we do, though, let's have a look at their construction example.

```{r error=TRUE}
URL <- "http://www.massey.ac.nz/~pscowper/ts/Maine.dat"
Maine.month <- read.table(URL, header=TRUE)
```

Data isn't available. Bummer. I even went to the university website and nothing doing. Grr.

OK, the important thing is that `frequency` is used to subdivide an interval into smaller increments which (probably?) relate to one another and form seasonal patterns. Probably.

Let's have another look at `aggregate` as that's how C&M convert a monthly time series into an annual one.
OK, having done that, I don't think that's the way to go. `aggregate` adds successive values together like `cumsum`, but the difference is that it pays attention to the `frequency`. When we aggregate the airline passenger data, it will collapse everything to one year. If we truly want quarterly data, we have to approach this differently.

```{r }
aggAP <- aggregate(AP)
class(aggAP)

length(aggAP)
length(AP)

(length(AP) / frequency(AP)) == length(aggAP)
```

There may be a `ts` way to do this, but I'm going to cheat and explore how to do that using `dplyr` or some such thing.

Tomorrow.

Note: altering a set of monthly figures to quarterly values means a few things. 1. The quarters may cross a calendar year. A rational quarter for "winter" would likely begin in December and end with February. This would mean summer is June, July and August, which makes sense. However, now quarters aren't assigned to proper years. Confusing. 2. If I'm not mistaken, we must collapse the months into a single observation. Either `sum` or `mean` would work. But I don't think we get the idea of three random observations from one quarter. I may have this all backwards, but I think there's some loss of information on the variability of the phenomena.

Back to chapter one. We've plotted, correlated and taken a window of observations. It's late and I'm not going to repeat that stuff. Something more profound happens on page 19. We decompose a time series and get an equation. That equation is number 1.2

$$ x_t = m_t + s_t + z_t$$

We know that $z$ will be stochastic. The parameters $m$ and $s$ look like deterministic parameters which we will estimate, though they vary at each point in time. We estimate m- the 'trend' by taking averages of the $x$'s around $x_t$. The seasonal parameter at time $t$ is then the difference between $x$ and $m$.

```{r }
decompAP <- decompose(AP)
class(decompAP)
plot(decompAP)

plot(stl(AP, "periodic"))
```

OK, so that's fairly cool. But how does this help me make forecasts and how do I assess the quality of my model? I might find that out in Chapter 2.
